Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-08-11 Manuscript/MPESubmission/Revision/response.tex

\raggedright
\setlength{\parindent}{0.5in}
\setcounter{secnumdepth}{0} 
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}

\setlength\parindent{0pt}

\begin{document}

% NC: Probably want to change figure to Figure, table to Table, appendix to Appendix to be consistent.

@@

\begin{enumerate}[(1)]
\item The Total Evidence method places living and fossil taxa on the same phylogeny.
\item Here we test the effect of missing morphological data on Total Evidence topologies.
\item The number of living taxa with morphological data has greatest effects on topology.
\item Bayesian inference outperforms Maximum Likelihood for correct clade recovery.
\item We recommend increased sampling effort for morphological data of living taxa.
\end{enumerate}

% -----------------
% Comment 4
% -----------------
\item{\textcolor{blue}{\textbf{Introduction:} Needs stronger emphasis on what specifically is unique about this study compared to what the many other studies on missing data have done and found.
Clearly outlining the questions in this study that are left open from previous research is important because some of the findings in this study suggest that there may be more error associated with missing data than previous research has suggested.}}

We have increased our emphasis on novelty throughout the introduction by more clearly stating the difference between the present study and the literature and more clearly outlining the question of this study, as follows:

% 0(before this section)-Missing data can be a problem
% 1-And until now, not many studies

% NC: Line numbers?
Until now, few attempts have been made to study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework (i.e. using both molecular and morphological data; e.g. Wiens et al., 2005; Manos et al., 2007; Pattinson et al., 2014).
% 2-These studies have done X and Y to see if missing data's a problem
These studies have assessed the effect of missing data on topology by either (1) comparing a dataset with missing data to subsets without missing data (Wiens et al. 2005); or (2) removing both molecular and some morphological data from living taxa to create artificial fossils (Manos et al., 2007; Pattinson et al., 2014).
% 3-They show that it's not a problem
Both approaches have shown that missing data are not a major problem and should not be an obstacle to combining both living and fossil species in the same phylogenies.
% 4-But they've done it a certain way
However, these studies were limited by their empirical-only approach and the use of only living taxa (Wiens et al., 2005) or the scheme of generating missing data (i.e. using the patterns from the fossil record; Manos et al., 2007; Pattinson et al., 2014).
% NC: This is still a problem (it's ok up to 4). WHY is an empirical-only approach limiting? WHY is it limiting to generate missing data using fossil patterns? 
Besides, the studies using fossil data, mainly focus on the palaeontological aspect of the question (i.e. the effect of missing data in fossils in Total Evidence matrices; Manos et al., 2007; Pattinson et al. 2014).
% NC: I don't like all these i.e.s here. Why not be specific in the text rather than throwing everything into brackets? If you're only citing manos and patterson then just say what they did.
% (next §)5-We're doing it this way

[lines 99-103 @@@] In this study, we propose a theoretical assessment of the effect of missing data in the Total Evidence method by removing living or fossil data, or all data for certain characters.
% NC: Could be more specific here and have a sentence that says. "This is an advance on previous studies because..."
We test the effect of missing data by measuring two crucial aspects of topology in both Maximum Likelihood and Bayesian phylogenies: (i) the conservation of clades and (ii) the displacement of wild-card taxa. 
% NC: Again how is this an advance. I guess it's related to this sentence from below:
% (cf. Wiens et al., 2005; Pattinson et al., 2014 measuring respectively only the clades support or the number of shared clades).

% TG: I actually rewrote this bit as well:
% NC: line numbers?
We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the resulting tree topology.
We inferred the topology from the matrices using both Maximum Likelihood and Bayesian Inference methods and measured the differences in topology using two different topological distance metrics as proxies for clade conservation (based on the Robinson-Foulds distance; Robinson and Foulds, 1981) and for wild-card taxa placement (based on the Triplets distance; Critchlow et al., 1996).
Thus, in the present study, we propose a new approach to the missing-data question, focusing on both palaeontological and neontological data (rather than just palaeontological data), using only simulated data and measuring the effects of missing data on two aspects of topology (cf. Wiens et al., 2005; Pattinson et al., 2014 measuring respectively only the clades support or the number of shared clades).
% NC: This paragraph seems to be repeating what you've already said above but in slightly more detail. Might make more sense to combine this paragraph and the one above it?

@@

% -----------------
% Comment 5
% -----------------
\item{\textcolor{blue}{\textbf{Abbreviation:} One suggestion is to consider abbreviating Maximum Likelihood and Bayesian Inference as ML and BI respectively early in the methods and use throughout.}}

We understand the reviewer's concern.
Nonetheless, we would prefer not abbreviate these terms. 
Although they might make sentences where Maximum Likelihood and/or Bayesian Inference are stated more concise, we think spelling out the terms makes our paper clearer and does not ask the reader to look back for the actual meaning of the abbreviation (Berlin 2013; Radiology; \url{http://pubs.rsna.org/doi/abs/10.1148/radiol.12121776}).
Additionally this could lead to a confusion between ML (Maximum Likelihood) and $M_L$ (Proportion of living taxa with no morphological data).
However, we can change this at a later stage if necessary.

% -----------------
% Comment 6
% -----------------
\item{\textcolor{blue}{\textbf{Methodology - True tree:} I suggest it warrants comparing at least the 'best' tree to the 'true' tree.}}

We have now compared both the ``best'' tree and the ``missing-data'' trees to the ``true'' tree as suggested by both reviewers.
The results are available in appendix B; figure B1 and appendix C; figures C6 and C7.
The ability to recover the ``true'' tree is poor for both methods and using both metrics (Maximum Likelihood 50\% CI between approximately 0.2 and 0.3 Normalised Robinson-Foulds and Triplets metric; Bayesian consensus tree 50\% CI between approximately 0.3 and 0.4 Normalised Robinson-Foulds metric and between 0 and 0.2 Normalised Triplets metric). % NC: What are these results comparing? All missing data trees  to the true tree? True trees to best trees? Clarify.
This may be related to the size of our matrices or to the ability of the algorithms to generate the ``complete matrix'' but also to our decision to put our minimum median bootstrap acceptance at 50 rather than higher values (line @@@ in the former manuscript and @@@ in our revision).
When comparing the ``true'' tree to the ``missing data'' trees, the results are \textit{scaled down} due to the low scores of the ``true'' \textit{vs.} ``best'' trees scores.
However, we can observe the same patterns between the methods as when using the ``best'' tree (i.e. Bayesian always outperforms ML in clade conservation but does poorly in wild-card taxa placement).
An interesting feature is that the Normalised Robinson-Foulds metric actually \textit{increases} with missing data % NC: specify what is being compared, e.g. "when we compare the missing data trees to the ``true'' trees."" 
This is because the difference between the ``best'' and the ``true'' tree is high, making trees with more data more different than those with lots of missing data.
When data are missing, however, more clades are unresolved and thus closer to clades from the ``true'' trees.

@@

Although though we agree with the comments of both reviewers that knowing the difference between the true and best/missing-data trees is important, we do not think that these results warrant inclusion in the main text.
In these simulations, we had to make a choice between (1) simulating larger matrices but with fewer chains (i.e. less variance) or (2) smaller matrices with more variance.
We deliberately made the second choice because our main question was "what is the effect of our parameters on recovering topology".
To answer this question, we maintain that comparing the best tree to the missing-data trees is the best approach for two reasons:
\begin{enumerate}
\item{The true tree is always unknown:} as we state in the manuscript: "In real world situations, the ``true'' tree is never available" (line 242 in the first manuscript and @@@ in the revision).
\item{The true tree is just the starting point of the simulations:} we use the true tree only as a starting point for our simulations.
Once generated, the true tree is fed to a cascade of algorithms and software (i.e. seqGen\{phyclust\}, rTraitDisc\{ape\}, RAxML, MrBayes) described in section 2.1 and 2.3 to get the best tree and the missing-data trees.
Comparing the latter to the true tree will answer our question "what is the effect of our missing-data parameters and the cascade of algorithms and software on recovering the true topology?".
\end{enumerate}
Therefore we added a second justification to the main text (see below this response) and added these analysis (true vs. best/missing-data trees) to the supplementary for readers that might be interested by this specific problem.

We added the following sentence to the main text to clarify and link to the supplementary analysis:

"We compare ``best'' trees to ``missing data'' trees but could also compare ``true'' trees to the ``missing data'' trees.
In practice, the difference between the ``best'' trees and the ``missing data'' trees represents the effect of our missing data parameters and of the phylogenetic methods used to infer the ``missing data'' trees.
The difference between the ``true'' and the ``missing data'' trees, however, represents the effect of our parameters used to generate the ``true'' tree and the algorithms used to generate the ``complete'' matrix as well as the effect of our missing data parameters and the phylogenetic methods used.
Because the main aim of this study is to look at the effect our missing data parameters on topological recovery, we chose to represent only the comparisons between the ``best'' trees and ``missing data'' trees.
The results of the comparisons of the ``true'' and the ``missing data'' tree are available in appendix B.
Note that this makes little difference to our overall results." line 276-280 @@@.

@@

See also appendix B section "Differences between the ``true'' and the ``best'' trees." and reviewer 2's comment 1 below.

% -----------------
% Comment 7
% -----------------
\item{\textcolor{blue}{\textbf{Methodology - Starting tree:} The authors should at least include some analysis in which no starting tree is given, the max. likelihood 'best' tree is given, and / or the 'best' tree is given but with some perturbation.
I fear that one of the main results may be biased due to this one methodological detail - a strong prior on tree topology.
The justification given was 'to speed up the Bayesian analysis' which does not seem adequate, given these datasets are small.}}

We took on board this comment and agree with the reviewer that if using the ``true'' tree as a starting tree had an effect it would bias our results.
We therefore ran an extra analysis to test if that was the case and found that it had no major effect of the inferred tree topologies.
We performed this extra analysis on a subset of five trees with the three missing data parameters at 0\%, 10\%, 25\%, 50\% and 75\% for 20 chains. 
We compared the topology of these trees in Bayesian inference when using the ``true'' tree as a starting tree or without using a starting tree (i.e. using the default in MrBayes which is to use a different random tree for each MCMC chain).
We found no significant difference in topologies between the Bayesian inferences starting with the true tree or a random tree.
Therefore, we do not believe our results are merely the result of using a specific starting tree topology.
These results are available in the supplementary materials Appendix A (Figure A2 and Table A1) and we also added a note to the manuscript:

"We used a fixed starting tree rather than a random starting tree (default MrBayes; Ronquist et al., 2012b) to speed up our Bayesian inferences.
Note that a starting tree is not a Bayesian prior on topology \textit{per se} and using a fixed starting tree did not significantly affect topology compared to using random starting trees (see Appendix A, section ``Effect of the starting tree on Bayesian inference'')." line 323-327 @@@.
See also Appendix A section "Effect of the starting tree on Bayesian inference".

@@

% -----------------
% Comment 8
% -----------------
\item{\textcolor{blue}{\textbf{Results - Bhattacharyya meaning:} Some descriptive statistics are given, but it would help to tell us if these differences are meaningful.
For example, on line 414 the Bhattacharrya Coefficients are given suggesting RF distances are lower for ML than BI trees at 0.69, 0.48, and 0.66 - but if this coefficient represents the probability of overlap between distributions then shouldn't the coefficients be $<$0.05 to say that one is lower than the other?
In the table, mean and median values of $<$0.01 and $>$0.85 are highlighted as being the most extreme.
Later the focus on Bhattacharyya Coefficients is 1 when distributions overlap completely and 0 when they do not - should this be the criterion?
Just make it clear early on so the readers knows what to think when they see these numbers. }}

We clarified the interpretation of the Bhattacharyya Coefficient throughout the manuscript.
We say that two distributions are significantly different when their probability of overlapping (the Bhattacharyya Coefficient) is $>$0.05 and that they are significantly similar when this probability is $>$0.95.
Even though other values (e.g. 0.85) are not conclusive, we still reported them to show the relative overlap of the distributions (i.e. two distributions with a coefficient of 0.85 are not significantly similar but are closer to each other than two distributions with a coefficient of 0.15; which are themselves not significantly different but still less close to each other than the previous ones).
We added the following statement to the main text and to the appendix B:

"The Bhattacharyya Coefficient is the probability of overlap between two distributions bounded between 0 (no overlap) and 1 (full overlap; Bhattacharyya, 1943, see Appendix B for calculation details)." lines @@@

"When the Bhattacharyya Coefficient between two distributions is $<$0.05, the distributions are significantly different.
When this coefficient is $>$0.95, the distributions are significantly similar.
Values between these two thresholds show the probability of overlap between the distributions but do not allow us to define the significance of the similarity or differences between distributions." line 419-424 @@@.

@@

% -----------------
% Comment 9
% -----------------
\item{\textcolor{blue}{\textbf{Results - Bayesian consensus vs. Bayesian posterior:} One major finding that was not fully addressed was that the distances between the BI posterior distributions of trees do not overlap the BI best or consensus trees.
It seems from the tables in Appendix C that the topology distance metrics for Bayesian posterior distribution are similar to those for the Bayesian consensus, albeit usually lower for RF distance than the Tr distance.
In Figure 4, it seems like the distributions of metrics based on CIs for Bayesian consensus and posterior overlap up until >25\% missing data.
How then in comparisons of the distributions are they always so low (Table 1 max prob. of overlapping distributions = 0.11)?}}

The difference between the Bayesian consensus and the Bayesian posterior trees can be due to two factors:
\begin{enumerate}[I]
\item the way we assess topological recovery is slightly different in each case: for the Bayesian consensus trees, we simply compared the missing-data consensus tree to the best consensus tree; for the Bayesian posterior trees however, we randomly compared 1000 missing-data Bayesian posterior trees to 1000 best Bayesian posterior trees.
Therefore, even if the comparison methodology is the same, the fact that the second methods draws random trees means we expect the topological recovery score to be blurred in comparison to other comparisons that do not involve random comparisons.
\item the Bayesian posterior trees are always resolved and thus, in the same way as the Maximum Likelihood tree, more likely to contain false-positive nodes. On the other hand, because the consensus tree is a strict majority consensus rule, only nodes with a a good support will be represented in the tree and all other nodes will be collapsed. This reduces the number of false-positive nodes (since an unresolved tree is always better than a wrongly resolved one) and increases clade conservation but also increases the displacement of wild-card taxa (since it is likely that some Triplets will just not exist in a poorly-resolved consensus tree).
\end{enumerate}
Both points were discussed in the original manuscript lines 566-573 (or lines 645-652 @@@ in the revised version).
We expanded the paragraph to:

@@

"Additionally, the Bayesian posterior trees performed more poorly than the Bayesian consensus tree (figure 4, table 1 and appendix C figure C5 and tables C5, C6 and C7).
This may be because the Bayesian posterior trees are always resolved and thus more likely to contain incorrectly resolved nodes (i.e. decreasing the Normalised Robinson-Foulds metric).
Conversely, the Bayesian consensus tree might not resolve nodes that are poorly supported and thus is more likely to contain only correctly resolved nodes (i.e. increasing the Normalised Robinson-Foulds metric)." lines 652-658. @@@

Concerning the last question of this comment: "How then in comparisons of the distributions are they always so low (Table 1 max prob. of overlapping distributions = 0.11)?", we apologise for the confusion here and assume it is due to our mislabeling of the figure 4 (see reviewer 1's comment 12).
In fact, when looking at all parameters separately (figure 4), there is no clear overlap between the Bayesian consensus trees and the Bayesian posterior trees (median Bhattacharyya Coefficient of 0.01; Table 1).
The maximum observed Bhattacharyya Coefficient (0.11; Table 1) is probably due to the combination of parameters: when all the parameters have many missing data, the probability of overlap between trees was slightly higher (0.11) but still really low.
For the Normalised Triplets distance however, the overlap is higher (median Bhattacharyya Coefficient of 0.59; table 1) as illustrated in figure 4 (i.e. much more overlap between the distributions).

### 2015-08-11 Manuscript/MPESubmission/Revision/response.tex

\documentclass[12pt,letterpaper]{article}

%%%% NC: FOR CLARITY - can you make the bits that are just copied from the text italic? Sometimes it jumps quickly between a comment to a piece of text. Italics will help make it clearer.%%%%

@@

% -----------------
% Comment 10
% -----------------
\item{\textcolor{blue}{\textbf{Results - Bayesian Triplets:} Another important result that is not mentioned is that using the wildcard taxon metric (Tr), Bayesian consensus trees do quite poorly with missing data.
Values in Figures 4 and 5 clearly drop below zero, although negative values are not illustrated on the axis, and this indicates that the tree topologies are no better than, or even worse than, random trees.
It is especially true when fossils are missing data, and that is going to be an important finding with practical application but must be addressed. }}

We modified the figures 4 and 5 to show Normalised Triplets metric values below 0 and added a dashed line at 0 to underline the difference in scale between both metrics.
As mentioned above (reviewer 1 comment 9), this is due to the fact that the Bayesian consensus tree will be less and less resolved when missing data increases.
This has two related effects.
The first one being to retain "good" clade conservation (i.e. avoiding false-positive nodes by collapsing unsupported ones).
And the second one is that when these clades are collapsed, it reduces the number of available triplets and thus increases the wild-card taxa score (Normalised Triplets metric).
In fact, when a clade ((a;b);c); is collapsed to (a;b;c), the whole clade is conserved (a, b and c are still in a common clade) however they also all become wild-card taxa: the position of a regarding b and c is now incorrect.

We added a note on this in the section about the effects of the tree inference method:

"The Bayesian consensus trees, however, perform badly for the Normalised Triplets metric: some parameter combinations, especially when the $M_F$ parameter reaches 75\% missing data, lead to negative values (figure 5).
A Normalised Triplets metric value below 0 means that the placement of some taxa is worse than expected by just randomly placing this taxon in the tree.
This can be interpreted as the absence of comparable triplets between some of the ``missing data" trees and ``best'' trees.
Even if clades are conserved (figure 5), the resolution within them can be poor to non-existent when a lot of data are missing (i.e. 75\%).
In such cases, the fossil taxa are likely to be placed in any of the clades that share the most characters.
These results are in agreement with previous studies that have showed that missing data can cause problems for recovering ``correct'' topologies, especially for small matrices of 100 characters (Wiens 2003).
It is important to note, however, that this effect can be reduced by increasing the number of characters (Wiens 2003)." lines 632-644. @@@

@@

% -----------------
% Comment 11
% -----------------
\item{\textcolor{blue}{\textbf{Conclusion:} The authors should clarify how their results compare to those of previous studies more directly to emphasize the nuances that make this study unique. }}

We rewrote the conclusion to deal with this comment (and the specific comments 13, 52 - see comment 52 for details on the choice of citing Arcila et al 2015. - and 56), as follows:

% TG: I rephrased as follows.
Previous studies have explored the effect of missing morphological data in Total Evidence matrices (Wiens et al., 2005; Manos et al.,2007, Pattinson et al., 2014).
The conclusions of theses studies, however, were limited by their empirical approach as well as several aspects of their experimental design (e.g. not including fossil taxa, Wiens et al., 2005; or focusing on the palaeontological aspect, Manos et al.,2007, Pattinson et al., 2014).
% NC: Again needs a bit more explanation. WHY is empirical approach a problem? Because it only gives answers to one specific scenario or clade?
Here we used a more in depth approach where missing data were generated from simulated data and according to three clearly defined missing-data parameters ($M_{L}$, $M_{F}$ or $N_{C}$).
This allowed us to confirm previous results that missing data can be especially problematic in small matrices (Wiens 2003), but also revealed the crucial importance of coding morphological data for living species in Total Evidence phylogenies.
Missing data in Total Evidence matrices is not a problem for recovering the ``best'' tree topology as long as enough living and fossil taxa in the matrix have data for overlapping morphological characters.
When missing data increases in any of our missing data parameters ($M_{L}$, $M_{F}$ or $N_{C}$), it reduces support for the placement of fossil taxa and increases the displacement of wildcard taxa.
Therefore we advise increased focus on coding morphological characters for a large number of the living taxa present in the matrix (i.e. 50\%) if the goal is to accurately combine both living and fossil species in phylogenies.
Doing so will increase overlap of morphological characters among living and fossil taxa, allowing the fossil taxa to be positioned relative to the living taxa based on their shared derived characters rather than simply on available data."

Additionally, the topology of the Bayesian consensus trees, regardless of the amount of missing data, were always closer to the ``best'' tree topology than the Maximum Likelihood trees.
This has also been observed in empirical data (e.g. Arcila et al., 2015) where Maximum Likelihood trees inferred from a Total Evidence matrix were less supported than the Bayesian consensus tree.
This might have an important impact on estimating topologies in the Total Evidence framework, because previous studies had to rely either on molecular scaffolds (e.g. Slater, 2013), taxonomic constraints (e.g. Slater, 2013; Beck and Lee, 2014) or even on fixing the topology (e.g. Ronquist et al., 2012a).
Therefore, we suggest extracting such topological backbones from the Bayesian consensus tree if needed.
% TG: maybe add in a non tip-dating analysis way (i.e. Nick Matzke comment)? 
% NC: Sure if it fits. If you want to "cite" Nick you can use (N.Matzke pers.comm.)

To conclude, the results of our analyses are encouraging and show that it is possible to accurately combine both neontological and palaeontological morphological data in the same phylogeny as long as there is sufficient overlap.
Hopefully, using these approaches will greatly improve our understanding of macroevolutionary patterns and processes." lines 602-634. @@@
% NC: I don't like this last bit. It's really weak. I'd cut it to be honest. The first sentence is not very clear (suddenly you revert back to neo and paleo) and the second sentence has no real explanation or link to the rest of the paper.

@@

% -----------------
% Comment 12
% -----------------
\item{\textcolor{blue}{\textbf{Figure 4:} Figure 4 has a color scheme issue, the colors in the caption do not match up to the figure.}}

We apologies for this silly mistake and thank the reviewer for pointing it out. We changed the colors to match with the paper's color scheme: "Maximum Likelihood trees (black), Bayesian consensus trees (grey), Maximum Likelihood bootstrap trees (blue) and Bayesian posterior tree distributions (orange)". See figure 4 caption.

@@

% -----------------
% Comment 13
% -----------------
\item{\textcolor{blue}{\textbf{Figure 4-5:} Am I wrong or shouldn't the y-axis for Normalised Triplets metric go below 0 - it looks those metrics for many missing data combinations have values below zero indicating trees are more different from the 'best' tree than expected by chance (Appendix B 1.1., see Appendix C Tables 1-3) and that is important to show.}}

We changed the axes of the graphs showing Normalised Triplets metric to show values below 0 throughout the manuscript and the appendices.

% -----------------
% Comment 14
% -----------------
\item{\textcolor{blue}{\textbf{Table caption:} Table caption should include a line about 'pooling across missing data schemes' to clarify what is presented.}}

We clarified the caption of the Table 1 as follows:

"Each line summarizes the probabilities of overlap between the distributions of the ``best'' tree versus trees from each inference method (Maximum Likelihood; Bayesian consensus; Maximum Likelihood Bootstraps and Bayesian posterior trees) pooled across all combinations of missing data parameter values, using the Normalised Robins-Foulds (RF) and Triplets (Tr) metrics. 
Values highlighted in bold are the extreme values of high or low probability of overlap between two methods. If two methods have a high probability of overlap, they have a similar ability to recover the ``correct'' tree topology." 
% NC: Do you want to say values > 0.95 are significantly similar, values < 0.05 are significantly different?

@@

% -----------------
% Comment 15
% -----------------
\item{\textcolor{blue}{\textbf{Previous research:} The missing-data problem has a long history in the literature and there are references to some of those older and recent papers.
But I feel it is necessary to summarize the following questions: what did those studies find, what are the questions those studies left open, and how does this study address them?}}

We now emphasize more of the recent findings from the missing-data papers in the introduction, and discussion (see comment 4 and 10 above).
We also included that our results for the $M_F$ parameter for the Normalised Robinson-Foulds metric were in accordance with Manos et al. 2007 in the discussion about the $M_F$ parameter:

"These results are in agreement with Manos et al. (2007) where as few as 16 characters were sufficient for correctly assigning artificial fossils to their correct clade." lines @@@

% NC: Try not to have too many " ".

@@

% -----------------
% Specific Comment 13
% -----------------
\item{\textcolor{blue}{\textbf{lines 80-103:} (1) the authors need to clearly justify the importance of this study compared the many studies that have investigated the effects of missing data on topology inference.
The justification currently given is not particularly strong and the specific unique questions asked in this study that are currently unanswered should be made clear.
The current argument does not always seem justified by the literature review.
For example, line 81 the authors argue that papers by Wiens and Pattinson examined morphological or molecular matrices separately, but in Pattinson, for example, combined datasets for living species were used to simulate extinct species by deleting the molecular and parts of the morphological datasets from living species.
The authors could argue that a limitation of the Pattinson study, which they address here, was the limited scheme for generating missing data from the datasets.
In this study, the authors perturbed the datasets more thoroughly in a clearly structured experimental design to investigate the effects of eliminating living versus fossil data or all data overall.
That is one aspect of this study that I perceived as new and unique.
(2) Another aspect of this study design that could be new but is not explored here is the effect of the missing data on the morphological substitution rate parameter estimates.
Can you recover the true values from the perturbed datasets?
This study also used a unique way of evaluating the tree inference under missing data scenarios.
These and other aspects of the study uniqueness must be clearer.}}

(1) - We thank the reviewer for this comment and now emphasize both the novel aspects of our study and the questions we are actually testing (see reviewer 1 comment 4 above).

(2) - the second point underlined by the reviewer in this specific comment strikes use as really interesting.
We ran this additional analysis by comparing the estimated $\alpha$ parameter for the morphological rates distribution to the ``true'' $\alpha$ parameter.
We found results that the $\alpha$ parameter is usually slightly underestimated for the various ``missing-data'' trees unless the number of living species with no morphological data is high ($M_L$ = 75\%), then, the $\alpha$ are overestimated.
These results are somewhat in agreement with our topological differences results: the number of living species seems to be a driving missing data parameter for inferring Total Evidence trees.
Nonetheless, we considered this study to already be long and complex.
We therefore think that adding an extra question to this study (i.e. what is the effect of missing data on the morphological substitution rate parameter estimates?) might lose the reader and make an already long paper even longer.
Therefore, we added the results of these analyses to \textbf{Appendix-A Morphological rates estimations results}.
We link to this analysis in the methods as follows:

@@

"The detailed MrBayes parameters are available in Appendix A along with details on the $\alpha$ parameter estimation." lines 336-338 @@@
See also Appendix A "Morphological rates estimations results".

% NC: Needs a bit more explanation. Something like, we also include in the appendix an analysis showing ... though it is beyond the scope of this paper to discuss these results further here.

@@

% -----------------
% Specific Comment 17
% -----------------
\item{\textcolor{blue}{\textbf{line 173:} streamline to "The substitution rates were selected from a gamma distribution with…" or something similar. }}

We modified the sentence as suggested.

% -----------------
% Specific Comment 18
% -----------------
\item{\textcolor{blue}{\textbf{line 173 to 176:} need some justification for why the substitution rate with alpha shape of 0.5 will avoid homoplasy. Yang 1996 is cited but clarify the justification of what is 'too much' homoplasy. }}

Since homoplasy is a difficult concept to measure (e.g. with character exhaustion; Wagner 2000, Evolution) it is difficult to assign a value for ``too much'' homoplasy.
We therefore removed this comment and changed the sentence to:
"In practice, a value of $\alpha$ $<$ 1 decreases the number of sites with high substitution rates, thus reducing homoplasic sites and increasing the phylogenetic signal (Hassanin et al., 1998; Estoup et al., 2002).%\citep{Hassanin1998611,EstoupHomoplasy}.
Also, we chose this $\alpha$ value to be consistent with our protocol for simulating morphological characters (see below)." (line 189-192@@@).

@@

% -----------------
% Specific Comment 21
% -----------------
\item{\textcolor{blue}{\textbf{line 185 to 188:} streamline, e.g. '…sampling with a probability of 0.85 for two states characters and 0.15 for three state characters (based on an empirical review of published matrices, see Appendix A).'}}

We modified the sentence as suggested.

% -----------------
% Specific Comment 22
% -----------------
\item{\textcolor{blue}{\textbf{line 193 to 196:} again, justify why gamma with shape of 0.5 is appropriate for minimizing homoplasy. This distribution places highest probability on substation rates less than 0.5, when one of the cited references potentially suggests that the lowest topological error was found at rates of ~0.5-1.5, and that very low substitution rates incurred error similar to very high rates ((Wright and Hillis 2014) 2014 Figure 3). Other cited references do not clearly state that this distribution and shape setting is optimal, so the authors really must justify. }}

We initially chose this parameter from personal communication with April Wright and to be consistent with the molecular rate parameter.
In figure 3 from Wright and Hillis 2014, even if the topological error is high when rates are near 0, as is mainly the case for parsimony and both parsimony and Mk with missing data, the fitted model for the Mk model with no missing data (light purple line) is still lowest when the rates are near 0.
Topological error, however, becomes minimal for all scenarios (Parsimony and Mk, with or without missing data) at a rate of 0.5 (see modified figure 3 from Wright and Hillis (2014) in this response; we added the tick marks and the 0.5 rate limit by using the ruler in Adobe Illustrator).

@@

\begin{figure}[]
\centering
    \includegraphics[keepaspectratio=true]{response_fig1.pdf}
\caption{Figure 3 from Wright and Hillis 2014 with added x axis ticks. The red line represents the 0.5 rate.}
\end{figure}

We do agree, however, that if rates get lower than 0.5, variance in topological error increases (especially for the Mk model with missing data).
Therefore, we implemented the \textit{a posteriori} matrix selection step to our analysis (i.e. keeping only the ``complete'' matrices that led to trees with a median bootstrap $>$ 50; explained lines 259-268 in the submitted manuscript and lines @@@ and @@@ in the revised one).

We rephrased the section highlighted by the reviewer as follows:

"We used low evolutionary rate parameters to be consistent with the molecular rate parameters, to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal (Wright and Hillis, 2014).
Topological error has been shown to be minimal at a morphological rate of 0.5 when using the M\textit{kv} model (Lewis, 2001; Wright and Hillis, 2014).
Note, however, that Wright and Hillis (2014) have shown that low morphological rates (< 0.5) increase variance in topological error, but we discarded simulations with such topological error by selecting only matrices with a “fair” phylogenetic signal (see Estimating phylogenies section below; Zander, 2004) so this should not influence our results." lines 211-217 @@@

@@

% -----------------
% Specific Comment 37
% -----------------
\item{\textcolor{blue}{\textbf{line 515:} '…the size of the matrix' missing 'of'}}

We fixed this typo.

% -----------------
% Specific Comment 38
% -----------------
\item{\textcolor{blue}{\textbf{line 518 to 520:} I am not sure that (Wagner 2000) states that incongruence between molecular and morphological data is 'more important in small morphological matrices'. }}

We now refer to two more classical citations: Bremer and Struwe 1992 (American Journal of Botany) and Patterson et al. 1993 (Annual Review of Ecology and Systematics), on phylogenetic signal and congruence between morphological and molecular data and also refer to a classical example of conflict probably due to having a small matrix (Masters and Brothers 2002; American Journal of Physical Anthropology; 36 characters) lines 578-579@@@.

@@

We modified the sentence as suggested.

% -----------------
% Specific Comment 43
% -----------------
\item{\textcolor{blue}{\textbf{lines 542-547:} These seem like the most important practical results and should be more strongly emphasized. These are very important results for practitioners and the authors should try to make it more clear to the reader that this is a key take-home message.}}

We modified the paragraph to:

"Surprisingly, the number of missing living taxa with morphological data ($M_{L}$) and the overall number of missing morphological characters ($N_{C}$), have a bigger effect than the amount of missing data for the fossil taxa ($M_{F}$).
For any additional missing living taxa with morphological data ($M_L$) beyond 50\%, there is no difference among trees with any combination of the other parameters ($M_F$ and $N_C$; Figure 6).
In other words, when the number of missing living taxa reaches 50\%, neither the amount of missing data in the fossil record ($M_F$), nor the number of characters in the matrix ($N_C$) affect topology.
A similar effect can be observed when the $N_C$ parameter reaches 50 characters (Figure 6).
This has important practical implications, especially for the best strategy to improve topology by collecting more morphological data (see below)." lines 589-609@@@

@@

% -----------------
% Specific Comment 45
% -----------------
\item{\textcolor{blue}{\textbf{line 557-558:} '…than the "best" maximum likelihood tree'}}

We modified the sentence as suggested.

% -----------------
% Specific Comment 46
% -----------------
\item{\textcolor{blue}{\textbf{line 549-565:} again, I am left wondering how much the differences in performance of max. likelihood and Bayesian inference techniques are driven by the use of the 'true' tree as a starting tree in the Bayesian MCMC search, while the likelihood analysis had no such guide tree to begin with. The authors at least need to make this point clear, if not re-run some subset of analyses without introducing a starting tree to test the effects of using the 'true' tree as a starting tree or not. I understand the computational limitations, but the practical limitation that empiricists will never have the 'true' tree to start with may (or may not) have a large impact on our interpretation and the generality of these findings. }}

As discussed above (reviewer 1, specific comment 29), we understand the reviewer's concern and have added an analysis in Appendix A (figure A2 and tables A1 and A2) where we tested the effect of using the ``true'' tree as a starting tree. We found that there is no significant effect of using the ``true'' tree instead of a random tree (MrBayes default) as a starting tree on topological recovery.

@@

We refer to this supplementary analysis in the main text as follows:

"Our results show that the topology of the Bayesian consensus tree is always closer to the ``best'' tree topology than the ``best'' Maximum Likelihood tree (Figure 5).
Note that the methodological choice of using the ``true'' tree as a starting tree for the Bayesian Inference rather than a random starting tree (see Methods), had no significant effect on topological recovery (see Appendix A, section ``Effect of the starting tree on Bayesian inference'' for details on the analysis of the effect of the starting tree in Bayesian inference)." lines 618-624@@@

@@

We also linked these results to previous studies (Wiens 2003; Systematic Biology) and discussed the negative values of the Normalised Triplets metric in the reviewer 1 comment 10 (see above).
